---
title: "HW 10"
author: "Ted Henson"
date: "4/17/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chapter 13, questions 1 and 3 (pp. 294-295), but omitting 1(h).

# Question 1
## a)
```{r}
library(faraway)
data(ohio)
length(unique(ohio$id))
nrow(unique(ohio[, c('id', 'smoke')]))
```

No, all mothers had the same smoking status throughout because all of the children's mothers had the same status for each child.

## b)
```{r}
library(ggplot2)
library(tidyverse)

ohio = ohio %>% group_by(id) %>% mutate(years = sum(resp))
sum.table = ohio %>% group_by(smoke,
                              years) %>% summarise(num.wheeze = n())
sum.table = sum.table %>% group_by(smoke) %>% mutate(prop.wheeze = num.wheeze / sum(num.wheeze))
sum.table
```

## c)
```{r}
ohio = left_join(ohio, sum.table)
smokers = ohio %>% dplyr::filter(smoke == 1)
non.smokers = ohio %>% dplyr::filter(smoke == 0)

ggplot(ohio, aes(prop.wheeze, years,
                 col = smoke)) + geom_point()

```

## d)
```{r}
ohio  = ohio %>% group_by(id) %>% mutate(total = sum(resp) / 4)
ohio = unique(ohio)
mod = glm(total ~ smoke, data = ohio, family = 'quasibinomial')
summary(mod)
```

This does not prove that there is a smoking effect because it could be due to random chance. The graph plotting years by smoking did not show any obvious difference.

## e)
```{r}
library(MASS)
modpq=glmmPQL(resp~age + smoke,random=~1|id,
               family=binomial, data=ohio)
summary(modpq)

```

Age has a significant effect whereas we do not have significant evidence by the p value to conclude that smoking has an effect.

## f)
```{r}
library(lme4)
modgh=glmer(resp~age + smoke + (1|id),
            nAGQ=25,family=binomial,ohio)
summary(modgh)


```

The fit and the p values were almost identical.


## g)
```{r}
library(INLA)

formula=resp~smoke +age +f(id,model='iid')
result=inla(formula,family='binomial',data=ohio)

# some of these variables are self-contained arrays or tables, for example

result$summary.fixed

result$summary.hyperpar

# others are complocated objects in their own right, for example

names(result$marginals.fixed)
```

The model shows that across the distribution age is negatively associated with resp and smoke has a high degree of variability so we cannot conclude that it is related to resp.

## i)
```{r}
library(geepack)
modgeep=geeglm(resp~ age + smoke,id=id,
               corstr='ar1',scale.fix=T,data=ohio,family=binomial)

summary(modgeep)

```

The model agrees with the previous models that smoking is not significant, but age is. The alpha of .5 with a small standard error shows that there is a high liklihood of autocorrelation within children: a child who already wheezes is likely to continue to wheeze.


# Question 3
## a)
```{r}
data(wavesolder)
head(wavesolder)
plot(y1+y2+y3 ~ ., data = wavesolder)
```
## b)
```{r}
mod1 = glm(y1 ~ prebake + flux + speed + preheat + cooling + agitator + temp, data = wavesolder,
           family = 'poisson')
summary(mod1)

mod2 = glm(y2 ~ prebake + flux + speed + preheat + cooling + agitator + temp, data = wavesolder,
           family 
           = 'poisson')
summary(mod2)
mod3 = glm(y3 ~ prebake + flux + speed + preheat + cooling + agitator + temp, data = wavesolder,
           family = 'poisson')
summary(mod3)
mod1 = glm(y1 ~ prebake + flux + speed + preheat + cooling + agitator + temp, data = wavesolder,
           family = 'quasipoisson')
summary(mod1)

mod2 = glm(y2 ~ prebake + flux + speed + preheat + cooling + agitator + temp, data = wavesolder,
           family 
           = 'quasipoisson')
summary(mod2)
mod3 = glm(y3 ~ prebake + flux + speed + preheat + cooling + agitator + temp, data = wavesolder,
           family = 'quasipoisson')
summary(mod3)
```
The first model is inadequate because of overdispersion. The second model includes an overdispersion paramater which eliminates the significance of many of the variables the intial model thought were significant.
## c)
```{r}

mod = glm(y1+y2+y3 ~ prebake + flux + speed + preheat + cooling + agitator + temp, data = wavesolder,
           family = 'quasipoisson')
summary(mod)

```

This model produced significant variables similar to the quasi poisson models for the individual runs.

## d)
```{r}
wavesolder$id = c(1:16)
wavesolder$sum = sum(wavesolder$y1,
                     wavesolder$y2,
                     wavesolder$y3)
wavesolder = gather(wavesolder,
           key = day,
           val = count,
           y1:y3)
wavesolder = wavesolder %>% arrange(id)

modgeep1=geeglm(count ~ prebake + flux + speed + preheat + cooling + agitator + temp,id = id,
               corstr='ar1',scale.fix=T,data=wavesolder,family=poisson)

summary(modgeep1)
```

With a fixed scale there is still no evidence that there is correlation between succesive duplicates.


## e)
```{r}
modgeep1=geeglm(count ~ prebake + flux + speed + preheat + cooling + agitator + temp,id = id,
               corstr='ar1',data=wavesolder,family=poisson)

summary(modgeep1)
```

Even without a fixed scale there is still no evidence that there is correlation between succesive duplicates.

## f)
```{r}
modgeep1=geeglm(count ~ prebake + flux + speed + preheat + cooling + agitator + temp,id = id,
               corstr='unstructed',data=wavesolder,family=poisson)

summary(modgeep1)
```

This model found a couple more significant variables than the poisson model did, and it produced lower p values for the significant variables.