---
title: "HW 7"
author: "Ted Henson"
date: "3/30/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup}
knitr::opts_chunk$set(message=F, warning=F,  eval = T)
```

# Question 2
## a)
```{r}
library(faraway)
data(coagulation)
head(coagulation)
library(ggplot2)
library(tidyverse)
ggplot(coagulation, aes(x = diet, y = coag)) + geom_boxplot()
```

It appears that diets D and A have low coagulation times comapred to diets B and C, which have very high coagulation times, especially C.

## b)
```{r}
op=options(contrasts=c('contr.sum','contr.poly'))
lmod=aov(coag~diet,coagulation)
summary(lmod)
new.data = data.frame(diet='D')
predict(lmod, new.data, interval = 'prediction')
```

## c)
```{r, message=F, warning=F}
library(lme4)
mmod=lmer(coag~diet+(1|diet),coagulation)
options(op)

summary(mmod)
new.data = data.frame(diet='D')
fit = predict(mmod, new.data,se.fit=T)

group.sd=as.data.frame(VarCorr(mmod))$sdcor[1]
resid.sd=as.data.frame(VarCorr(mmod))$sdcor[2]
pv=numeric(1000)
for(i in 1:1000){
y=unlist(simulate(mmod))
bmod=suppressMessages(refit(mmod,y))
pv[i]=predict(bmod,new.data=data.frame(operator='D'))+rnorm(n=1,sd=resid.sd)
}
quantile(pv,c(0.025,0.975))
```

## d)
```{r}

for(i in 1:1000){
y=unlist(simulate(mmod))
bmod=suppressMessages(refit(mmod,y))
pv[i]=predict(bmod,re.form=~0)[1]+rnorm(n=1,sd=group.sd)+rnorm(n=1,sd=resid.sd)
}
quantile(pv,c(0.025,0.975))
```

## e)
```{r}

for(i in 1:1000){
y=unlist(simulate(mmod))
bmod=suppressMessages(refit(mmod,y))
pv[i]=predict(bmod,re.form=~0)[1]+rnorm(n=1,sd=group.sd)+rnorm(n=1,sd=resid.sd)
}
quantile(pv,c(0.025,0.975))
```

# Question 3
## a)
```{r}
data(eggprod)
ggplot(eggprod, aes(x=block, y = eggs)) + geom_boxplot()
ggplot(eggprod, aes(x=treat, y = eggs)) + geom_boxplot()

ggplot(eggprod, aes(x=treat, y = eggs)) + geom_point() + facet_grid(~ block)

```

Across the board treatment o and block 4 appear to produce the lowest eggs, but when combined, these effects do not remain. The lowest number of eggs occured with treatment O and block 2.

## b)
```{r}
lmod=aov(eggs~treat+block,eggprod)
summary(lmod)

```

Based on the p values the treatment variable apperas to be significant, but the block variable may not be after accounting for the treatment.

## c)
```{r}
mmod=lmer(eggs~treat+(1|block),eggprod)
summary(mmod)
fixef(mmod)


```

According to this model, treatment E is best at maximizing egg production, but we cannot be sure because of the random effects imposed by the block variable.

## d)
```{r}
library(pbkrtest)
nmod=lmer(eggs~1+(1|block),eggprod)

KRmodcomp(mmod,nmod)


```

The Kenward-Roger approximation agrees with the fixed effects model that the treatments are significant.

## e)
```{r}
#parametric bootstrap
amod=lmer(eggs~treat+(1|block),eggprod,REML=F)
nmod=lmer(eggs~1+(1|block),eggprod,REML=F)
pmod=PBmodcomp(amod,nmod)
summary(pmod)


```

The paramatric bootstrap agrees that the treatments are significant.

## f)
```{r}
rmod=lmer(eggs~treat+(1|block),eggprod)
nlmod=lm(eggs~treat,eggprod)
as.numeric(2*(logLik(rmod)-logLik(nlmod,REML=T)))

lrtstatf=numeric(1000)
for(i in 1:1000){
reggs=unlist(simulate(nlmod))
nlmodr=lm(reggs~treat,eggprod)
#rmodr=refit(rmod,reggs)
rmodr=lmer(reggs~treat+(1|block),eggprod)
lrtstatf[i]=2*(logLik(rmodr)-logLik(nlmodr,REML=T))
}

mean(lrtstatf<0.00001)
mean(lrtstatf>as.numeric(2*(logLik(rmod)-logLik(nlmod,REML=T))))
```

The bootstrapping test agrees with the fixed results test that the blocks are not significant.
