---
title: "STOR 455 Assignment 2"
author: "Ted Henson"
date: "September 9, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tibble)
library(readr)
```

```{r}
setwd("~/STOR 455")
UsedCarLot <- read_csv("UsedCarLot.csv")
honda <- UsedCarLot[which(UsedCarLot$make == 'Honda'),]
accord <- honda[which(honda$model == 'Accord'),]



```
##Exercises 1 & 2

```{r}
cardata <- accord
cardata$age <- 2018 - cardata$year
linear <- lm(price~age, data = cardata)
plot(x = cardata$age, y = cardata$price)
abline(linear)
summary(linear)
```

The slope estimate of -2.2085 means that for every year the honda accord ages, it's value decreases by $2,2085. The negative slope makes sense because a car's price will decrease with age.


##Exercise 3

```{r}
qqnorm(linear$residuals)
qqline(linear$residuals)
hist(linear$residuals)

```


The data appears to be mostly linear, although there is a slight skew in the upper quantile. Our model overprices cars that are very old.


##Exercise 4

```{r}
outlier <- which(linear$residuals == max(abs(linear$residuals)))
print(outlier)
students <- rstudent(linear)
standard <- rstandard(linear)
students[outlier]
standard[outlier]
```

Both student and starndized residuals are more than 3 standard deviations away from the mean so the data point should probably be classified as an outlier.



##Exercise 5

```{r}
interval <- confint(linear, level = .90)
interval
```
We can say with 90% confidence that the slope of the best fine line of these variables is between the interval (-2.45566, -1.961407).



## Exercise 6

```{r}

cor.test(cardata$price, cardata$age)

summary(linear)


anova(linear)

```


## Exercise 7

```{r}

#subset for just 5 year old cars?
theanswer <- linear$coefficients[2]*5 + linear$coefficients[1]
theanswer

```


Our model predicts that a five year old Honda Accord will cost $29,0895. 


```{r}
whoknows <- data.frame(age=5)
predict.lm(linear, whoknows, interval = 'confidence')

```



If repeated samples were taken and 90% confidence intervals were computed, 90% of the intervals computed would contain the mean price at this age.

```{r}
predict.lm(linear, whoknows, interval = 'prediction')

```


If repeated samples were taken and 90% confidence intervals were computed, 90% of the intervals computed would contain the price of an individual car at this age.


##Exercise 8

```{r}

y  = linear$coefficients[1] / -linear$coefficients[2]
y


```

According to the model a car over 18.17141 years of age should be free. This phenomenon shows that our model does not do a good job of estimating prices for cars that are very old, because although an 18.17 year old car would be very cheap, it would not be free.


##Exercise 9
```{r}

linear <- lm(log(price)~log(age), data = cardata)
plot(x = log(cardata$age), y = log(cardata$price))
abline(linear)
summary(linear)

```

This model has a lower R squared than the standard regression. This model may be better for very new and very old cars, but on average this model performs worse.
