---
title: "HW 1"
author: "Ted Henson"
date: "August 29, 2018"
output:
html_document: default
html_notebook: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

### Exercise 1

Part A


```{r}
library(Stat2Data)
library(tibble)
data(HighPeaks)
#load("C:/Users/tedhe/OneDrive/Documents/R/win-library/3.5/Stat2Data/data/HighPeaks.rda")
HighPeaks <- as_tibble(HighPeaks)

leastsquares <- lm(Time ~ Ascent, data = HighPeaks)
print(paste("Least Squares line: ", 'Time = ',  leastsquares$coefficients[2], 'Ascent', ' + ', leastsquares$coefficients[1], sep = ''))

```

Part B


```{r}
plot(x = HighPeaks$Ascent, y = HighPeaks$Time, xlab = 'Ascent', ylab = 'Time', main = 'Plot Of Ascent Versus Time')
abline(leastsquares)
```


The plot looks somewhat linear, however there is a decent amount of variation.


Part C
```{r}
HighPeaks$Peak <- gsub(' ', '', as.character(HighPeaks$Peak), fixed = T)

tooearlyfothis <- HighPeaks[which(HighPeaks$Peak == "Mt.Marcy"), 'Ascent']

theprediction <- leastsquares$coefficients[2] * tooearlyfothis + leastsquares$coefficients[1]
print(paste(theprediction, 'hours is the time predicted for Mt. Marcy'))

theresidual <-  HighPeaks[which(HighPeaks$Peak == "Mt.Marcy"), 'Time'] - theprediction
print(paste(theresidual, 'is the residual for Mt. Marcy'))


```


Part D
```{r}
theworst <- max(abs(leastsquares$residuals))

index <- which(leastsquares$residuals == theworst)

themountain <- HighPeaks[index, 'Peak']
print(themountain)
print(theworst)

```


Part E

```{r}
plot(y = leastsquares$residuals, x = leastsquares$fitted.values, xlab = 'Predictions', ylab = 'Residuals', main = 'Plot Of Predictions versus Residuals')
abline(a= 0, b = 0)


```



The distribution of errors appears to be centered around zero for the residual plot, showing zero mean.





```{r}
qqnorm(leastsquares$residuals)
qqline(leastsquares$residuals)
```



Linearity may not be satisfied because there may be some relationship between the errors on the lower quantile.





```{r}
hist(leastsquares$residuals)

```





There is a skew right, so perhaps our errors are not independent as shown by the histogram and normal quantile plot.



### Exercise 2


Part A
```{r}
data(Perch)
plot(x = Perch$Length, y = Perch$Weight, xlab = 'Length', ylab = 'Weight', main = 'Plot of Length Versus Weight')

```


The plot looks exponential rather than linear.

Part B
```{r}
model <- lm(log(Perch$Length) ~ log(Perch$Weight))
summary(model)


```



```{r}
ggplot(Perch, mapping = aes(log(Length), log(Weight))) + geom_point() + xlab('Length') + ylab ('Weight') +  ggtitle('Plot Of length Versus Weight') + 
geom_smooth()

```


This plot looks linear after the logistic transformation.

Part C



```{r}

qqnorm(model$residuals)
qqline(model$residuals)
hist(model$residuals)

```



There appears to be some relationship between the errors in the lower quantile as shown by the histogram and qqnorm plot.




Part D
```{r}
expectedweight <- 20*model$coefficients[2] + model$coefficients[1]
expectedweight

```

